### 2015-07-01 doc/thesis/system.tex
Minor fixes to the section about table refinement

First, our type system uses the auxiliary relation
$\env_{1}, \penv \vdash [e_{1}] = e_{2} : (F,V), \env_{2}$ to type check each
table field.
This auxiliary relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, checking a table field $[e_{1}] = e_{2}$
produces a pair $(F,V)$ and a new type environment $\env_{2}$.
A pair $(F,V)$ means that $e_{1}$ has type $F$ and $e_{2}$ has type $V$,
where $F$ is the type of the key and $V$ is the type of the value.

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
We made this restriction because our type system does not keep
track of aliases to table fields.
This means that allowing \emph{unique} and \emph{open} table
types to appear in the type of a value would allow the
creation of unsafe aliases.
The rules that type check table fields use the auxiliary function
\emph{fix} to change any \emph{unique} and \emph{open} table types
used in the field initializer to \emph{fixed}.

The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
table constructor with a non-empty list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(F_{i},V_{i}), \env_{i+1} \\
       T = \{F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
       \env_{f} = merge(\env_{1}, ..., \env_{n+1})
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
\end{array}
\]

Back to our example, the constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR2}.

@@

The rule \textsc{T-REFINE1} uses the auxiliary function \emph{vt}
to adjust the type of the value according to the type of the key.
More precisely, \emph{vt} includes the type $\Nil$ in the type of
the value when the type of the key is not a literal type.
It is also this function that uses \emph{fix} to prevent
\emph{unique} and \emph{open} table types to appear in the type
of the values.

@@

Note that the coercion rules only allows changing the type
of a variable if the new type is a supertype of the previous type,
and the resulting type is always \emph{fixed} or \emph{closed}
to prevent the creation of unsafe aliases.

@@

Rule \textsc{T-FUNCTION1} illustrates this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION1}\\
\dfrac{\begin{array}{c}
       closeall(\env_{1})[\overline{id \mapsto F}], \penv[\ret \mapsto S] \vdash s, \env_{2} \\
       \env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s)) \\
       \env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s))
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_{1} \times ... \times F_{n} \times \Nil{*} \rightarrow S, \env_{4}}
\end{array}
\]

@@

>>>>

First, Typed Lua uses the auxiliary relation
$\env_{1}, \penv \vdash [e_{1}] = e_{2} : (F,V), \env_{2}$ to type check each
table field.
This auxiliary relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, checking a table field $[e_{1}] = e_{2}$
produces a pair $(F,V)$ and a new type environment $\env_{2}$.
A pair $(F,V)$ means that $e_{1}$ has type $F$ and $e_{2}$ has type $V$,
where $F$ is the type of the key and $V$ is the type of the value.

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
We made this restriction because our type system does not keep
track of aliases to table fields.
This means that allowing \emph{unique} and \emph{open} table
types to appear in the type of a value would allow the
creation of unsafe aliases.
Due to this restriction, the rule that type check table fields
use the auxiliary function \emph{fix} (in the definition of \emph{vt})
to change any \emph{unique} and \emph{open} table types
used in the field initializer to \emph{fixed}.
Rule \textsc{T-FIELD} defines this behavior:
\[
\begin{array}{c}
\mylabel{T-FIELD}\\
\dfrac{\env_{1}, \penv \vdash e_{2}:V, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{1}:F, \env_{3}}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (F,vt(V)), \env_{3}}
\end{array}
\]

The rule \textsc{T-FIELD} uses the auxiliary function \emph{vt}
to adjust the type of the value according to the type of the key.
More precisely, \emph{vt} includes the type $\Nil$ in the type of
the value when the type of the key is not a literal type.
It also uses \emph{fix} to prevent \emph{unique} and \emph{open}
table types to appear in the type of the values.
We define these functions as follows:
\begin{align*}
vt(L, V) & = fix(V) \\
vt(F_1, F_2) & = nil(fix(F_2))\\
vt(F_1, {\bf const}\;F_2) & = {\bf const}\;nil(fix(F_2))\\
\\
nil(T) & = T \;\;\; \mathrm{if}\;\Nil \lesssim T\\
nil(T) & = T \cup \Nil \;\;\; \mathrm{otherwise}\\
\\
fix(F_{1} \cup F_{2}) & = fix(F_{1}) \cup fix(F_{2})\\
fix(\{\overline{F{:}V}\}_{unique|open}) & = \{\overline{F{:}V}\}_{fixed} \\
fix(F) & = F
\end{align*}

The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
table constructor with a list of table fields that do not end with
an expression that potentially returns multiple values:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(F_{i},V_{i}), \env_{i+1} \\
       T = \{F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
       \env_{f} = merge(\env_{1}, ..., \env_{n+1})
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
\end{array}
\]

Back to our example, the constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR2}.
The subtyping rule for \emph{unique} table types allows us assigning
this table to a variable with a more general type such as
$\{1:\String, 2:\String, 3:\String\}_{closed}$ or even
$\{\Integer:\String \cup \Nil\}_{closed}$.

@@

The rule for refining \emph{open} table types is similar,
changing only the tag in the type of $id$:
\[
\begin{array}{c}
\mylabel{T-REFINE2}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \{\overline{F{:}V}\}_{open}\\
       \env_{1}, \penv \vdash e:F_{new}, \env_{2} \;\;\;
       \nexists i \in 1..n \; F_{new} \lesssim F_{i} \;\;\;
       V_{new} = vt(F_{new},V) \;\;\; n = |\overline{F{:}V}|
       \end{array}}
      {\env_{1}, \penv \vdash id[e] {<}V{>}:V_{new}, \env_{2}[id \mapsto \{\overline{F{:}V}, F_{new}{:}V_{new}\}_{open}]}
\end{array}
\]

Our type system also includes analogous rules for adding methods
to \emph{unique} and \emph{open} tables as a side-effect of
type checking a method declaration, but we will not discuss them
in this section for brevity.

@@

Note that the coercion rules only allows changing the type
of a variable if the new type is a supertype of the previous type,
and the resulting type is always \emph{fixed} or \emph{closed}
to prevent the creation of unsafe aliases.
If the coercion is to a \emph{closed} table type the type of the
table changes to an \emph{open} table type with the same shape,
but if the coercion is to a \emph{fixed} table type the table
has to assume the same type.

@@

Rule \textsc{T-FUNCTION1} types non-variadic function declarations,
and it illustrates this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION1}\\
\dfrac{\begin{array}{c}
       closeall(\env_{1})[\overline{id \mapsto F}], \penv[\ret \mapsto S] \vdash s, \env_{2} \\
       \env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s)) \\
       \env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}F}){:}S \; s))
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_{1} \times ... \times F_{n} \times \Nil{*} \rightarrow S, \env_{4}}
\end{array}
\]

@@

%The rules for declaring variadic functions and recursive functions
%are similar to \textsc{T-FUNCTION1}, and we will not
%show them in this section for brevity.
